Gaussian Processes Autoregressive Models for Forecasting the Disturbance Storm Time Index

### 2016-05-23 swsc.tex
Updates from ShareLaTeX

Our aim is to infer the function $f(\mathbf{x})$ from the noisy training data and generate predictions $f(\mathbf{x}^{*}_i)$ for a set of test points $ {\mathbf{x}^{*}_i : \forall i \in 1, \cdots, M} $. We define $\mathbf{X}^*$ as the test data matrix whose rows are formed by $\mathbf{x}^{*}_i$ as shown in equation \ref{eq:testfeat}. Using the multivariate Gaussian distribution in equation \ref{eq:sto} we can construct the distribution of $f(\mathbf{x})$ over the training and test points. The vector of training and test outputs $\left( \begin{array}{c} \mathbf{y} \\ \mathbf{f_*} \end{array} \right)$ is of dimension $(N+M) \times 1$ and is constructed by appending the test set predictions $\mathbf{f}_* = \left( \begin{array}{c} f(\mathbf{x^{*}}_1) \\ f(\mathbf{x^{*}}_2) \\ \vdots \\ f(\mathbf{x^{*}}_M) \end{array} \right)$ to the observed noisy measurements $\mathbf{y}$.

@@

\citet{Ji2012} compare six OSA $Dst$ models in their paper by compiling a list of 63 geomagnetic storm events of varying strengths which occurred from 1998-2006. This serves as an important stepping stone for a systematic comparison of existing and new prediction techniques because performance metrics are averaged over a large number of storm events that occurred over a 8 year period. They compare the averaged performance of these models on four key performance metrics.

@@

Table \ref{table:DstModels} gives a brief enumeration of the models compared, we use the performance benchmark results published in \citet{Ji2012} and to them add results obtained from testing \emph{GP-AR}, \emph{GP-ARX} as well as the \emph{Persistence} model on the same list of storms.

For the purpose of generating the cumulative probability plots of the model errors, we must obtain hourly predictions for all the storm events using the \emph{NARMAX} and \emph{TL} models. In the case of the \emph{NM} model, the formula outlined in \citet{balikhin:narmax} is used to generate hourly predictions. For the \emph{TL} model we use real time predictions listed on their website (\url{http://lasp.colorado.edu/home/spaceweather/}) corresponding to all the storm events. The real time predictions are at a frequency of ten minutes, hourly predictions are generated by averaging the $Dst$ predictions for each hour.

@@

\begin{figure}
   \centering
   \includegraphics[width=\textwidth]{Compare_CC.png}
      \caption{Averaged cross correlation performance of GP-AR, GP-ARX and the Persistence model versus results in \citet{Ji2012}.}
         \label{fig:cc}
   \end{figure}

With respect to the \emph{TL} model one important concern must be expressed. The training of the \emph{TL} model was done on Omni data from 1996-2002 which overlaps with a large number of the storm events, therefore the experimental procedure has a strong bias towards \emph{TL} model.

\begin{figure}
   \centering
   \includegraphics[width=\textwidth]{Compare_deltaDst.png}
      \caption{Averaged $\Delta Dst_{min}$ of GP-AR, GP-ARX and the Persistence model versus results in \citet{Ji2012}.}
         \label{fig:deltaDst}
   \end{figure}

@@

Figures \ref{fig:rmse}, \ref{fig:cc}, \ref{fig:deltaDst} and \ref{fig:timingErr} compare the performance of \emph{GP-AR}, \emph{GP-ARX} and the \emph{Persistence} model with the existing results of \citet{Ji2012}. Figure \ref{fig:rmse} compares the average RMSE of the models over the list of 63 storms, we can see that the \emph{GP-ARX} model gives an improvement of approximately $38\%$ in RMSE with respect to the \emph{TL} model and $60\%$ improvement with respect to the \emph{NM} model.

Figure \ref{fig:relprob} compares the empirical distributions of absolute relative model errors across all storm events in the experiments. If one chooses a $95\%$ probability bound then it can be seen that the \emph{GP-ARX} model bounds the relative error to $20\%$ of the actual $Dst$ value, while in the case of the \emph{NM} model it bounds the relative error to $50\%$ of the actual $Dst$ value. Therefore the \emph{GP-ARX} model gives an approximate improvement of $25\%$ over the other compared models on the test data set chosen.

\begin{figure}
   \centering
   \includegraphics[width=\textwidth]{Compare_timingerr.png}
      \caption{Averaged absolute timing error of GP-AR, GP-ARX and the Persistence model versus results in \citet{Ji2012}.}
         \label{fig:timingErr}
   \end{figure}

Figure \ref{fig:cc} shows the comparison of correlation coefficients between model predictions and actual $Dst$ values. From the results of \citet{Ji2012}, the \emph{TL} model gives the highest correlation coefficient of $94\%$, but that is not surprising considering the fact that there is a large overlap between the training data used to train it and storm events. This is also the case in \ref{fig:relprob} where we average predictions generated from a pre-trained instance of the \emph{TL} model. Even so the \emph{GP-ARX} model gives comparable correlation coefficient to \emph{TL} a testimony to its performance given that its training set has no overlap with the storm events.

\begin{figure}
   \centering
   \includegraphics[width=\textwidth]{Compare_pred.png}
      \caption{Comparison of OSA predictions generated by the NM, TL and GP-ARX models for the storm event $9^{th}$ November $2004$ 11:00 UTC - $11^{th}$ November $2004$ 09:00 UTC}
         \label{fig:predictions}
\end{figure}

@@

Figure \ref{fig:timingErr} compares the time discrepancy of predicting the storm peak, again as observed above the \emph{Persistence} model will always have $\Delta t_{peak} = -1$, the GP-AR and GP-ARX models give better performance than the other models in terms of timing error in prediction of storm peaks.

In figure \ref{fig:predictions} we show the OSA predictions generated by the TL, NM and GP-ARX models for one event from the list of storms in \citet{Ji2012}. This storm is interesting due to its strength ($Dst_{min} = -289 nT$) and the presence of two distinct peaks. In this particular event the GP-ARX model approximates the twin peaks as well as the onset and decay phases quite faithfully which is lacking in the NM and TL model predictions. The TL model recognises the presence of twin peaks but over estimates the second one as well as the decay phase, while the NM model is much delayed in the prediction of the initial peak and fails to approximate the time and intensity of the second peak.

As noted in section \ref{sec:inference}, \emph{Gaussian Process} models generate probabilistic predictions instead of point predictions, giving the modeller the ability to specify error bars on the predictions. In figure \ref{fig:predictionswitherrorbar} OSA predictions along with lower and upper error bars generated by the GP-ARX model are depicted for the event used as a reference in figure \ref{fig:predictions}. The error bars are calculated at one standard deviation on either side of the mean prediction, using the posterior predictive distribution in equations \ref{eq:posteriormean} and \ref{eq:posteriorcov}.

\begin{figure}
   \centering
   \includegraphics[width=\textwidth]{Compare_pred_err_bar.png}
      \caption{OSA predictions with error bars $\pm 1$ standard deviation, generated by the GP-ARX model for the storm event $9^{th}$ November $2004$ 11:00 UTC - $11^{th}$ November $2004$ 09:00 UTC}
         \label{fig:predictionswitherrorbar}
\end{figure}

@@

We conclude this paper with three key conclusions/insights learned from our research so far.
   \begin{enumerate}
      \item \emph{Persistence} model must be central in the model evaluation process in the context of \emph{One Step Ahead} prediction of the $Dst$ index. It is clear that the \emph{persistence} behavior in the $Dst$ values is very strong i.e. the trivial predictive model $\hat{Dst}(t) = Dst(t-1)$ gives excellent performance. In fact it can be seen in \ref{sec:res} that the \emph{Persistence} model is outperforms every model compared by \citet{Ji2012} in the context of OSA prediction of $Dst$. Therefore any model that is proposed to tackle the OSA prediction problem for $Dst$ must be compared to the \emph{Persistence} model and show visible gains above it.

@@

      If one compares the training sets used by the models, one can appreciate that they are able to learn robust models for prediction with relatively small training and validations sets (training set contains 250 instances, while the validation set contains 432 instances) on the other hand the \emph{NM} model uses one year (8760 instances) of data to learn the formula outlined in \citet{balikhin:narmax} and the \emph{TL} model uses six years of data (52560 instances) for training.

      The encouraging results of Gaussian Processes illustrate the strengths of the Bayesian approach to predictive modeling. Since the GP models generate predictive distributions for test data and not just point predictions they lend themselves to the requirements of space weather prediction very well because of the need to generate error bars on predictions.

      \item Multi-step ahead prediction should be the preferred gold standard with respect to forecasting of geomagnetic indices. Also important is the usage of more informative performance metrics for evaluating and comparing empirical predictive models.
   \end{enumerate}

>>>>

Our aim is to infer the function $f(\mathbf{x})$ from the noisy training data and generate predictions $f(\mathbf{x}^{*}_i)$ for a set of test points $ {\mathbf{x}^{*}_i : \forall i \in 1, \cdots, M} $. We define $\mathbf{X}^*$ as the test data matrix whose rows are formed by $\mathbf{x}^{*}_i$ as shown in equation \ref{eq:testfeat}. Using the multivariate Gaussian distribution in equation \ref{eq:sto} we can construct the distribution of $f(\mathbf{x})$ over the training and test points. The vector of training and test outputs $\left( \begin{array}{c} \mathbf{y} \\ \mathbf{f_*} \end{array} \right)$ is of dimension $(N+M) \times 1$ and is constructed by appending the test set predictions $\mathbf{f}_*$ to the observed noisy measurements $\mathbf{y}$.

@@

\citet{Ji2012} compare six \emph{one step ahead} $Dst$ models as listed in table \ref{table:DstModels} in their paper by compiling a list of 63 geomagnetic storm events of varying strengths which occurred from 1998-2006. This serves as an important stepping stone for a systematic comparison of existing and new prediction techniques because performance metrics are averaged over a large number of storm events that occurred over a 8 year period. They compare the averaged performance of these models on four key performance metrics.

@@

Table \ref{table:DstModels} gives a brief enumeration of the models compared. We use the same performance benchmark results published in \citet{Ji2012} and we add the results obtained from testing \emph{GP-AR}, \emph{GP-ARX} as well as the \emph{Persistence} model on the same list of storms.

For the purpose of generating the cumulative probability plots of the model errors, we need to obtain hourly predictions for all the storm events using the \emph{NARMAX} and \emph{TL} models. In the case of the \emph{NM} model, the formula outlined in \citet{balikhin:narmax} is used to generate hourly predictions. For the \emph{TL} model we use real time predictions listed on their website (\url{http://lasp.colorado.edu/home/spaceweather/}) corresponding to all the storm events. The real time predictions are at a frequency of ten minutes and hourly predictions are generated by averaging the $Dst$ predictions for each hour.

@@

\begin{figure}
   \centering
   \includegraphics[width=\textwidth]{Compare_CC.png}
      \caption{Averaged cross correlation performance of \emph{GP-AR}, \emph{GP-ARX} and the \emph{Persistence} model versus results in \citet{Ji2012}.}
         \label{fig:cc}
\end{figure}

With respect to the \emph{TL} model one important caveat must be noted. The training of the \emph{TL} model was done on Omni data from 1996-2002 which overlaps with a large number of the storm events, therefore the experimental procedure has a strong bias towards \emph{TL} model.

@@

\begin{figure}
   \centering
   \includegraphics[width=\textwidth]{Compare_deltaDst.png}
      \caption{Averaged $\Delta Dst_{min}$ of \emph{GP-AR}, \emph{GP-ARX} and the \emph{Persistence} model versus results in \citet{Ji2012}.}
         \label{fig:deltaDst}
   \end{figure}

Figures \ref{fig:rmse}, \ref{fig:cc}, \ref{fig:deltaDst} and \ref{fig:timingErr} compare the performance of \emph{GP-AR}, \emph{GP-ARX} and the \emph{Persistence} model with the existing results of \citet{Ji2012}. Figure \ref{fig:rmse} compares the average RMSE of the predictions over the list of 63 storms. One can see that the \emph{GP-ARX} model gives an improvement of approximately $38\%$ in RMSE with respect to the \emph{TL} model and $60\%$ improvement with respect to the \emph{NM} model. It can also be seen that the \emph{Persistence} model gives better \emph{RMSE} performance than the models compared in \citet{Ji2012}.

\begin{figure}
   \centering
   \includegraphics[width=\textwidth]{Compare_timingerr.png}
      \caption{Averaged absolute timing error of \emph{GP-AR}, \emph{GP-ARX} and the \emph{Persistence} model versus results in \citet{Ji2012}.}
         \label{fig:timingErr}
   \end{figure}

Figure \ref{fig:cc} shows the comparison of correlation coefficients between model predictions and actual $Dst$ values. From the results of \citet{Ji2012}, the \emph{TL} model gives the highest correlation coefficient of $94\%$, but that is not surprising considering the fact that there is a large overlap between the training data used to train it and storm events. This is also the case in \ref{fig:relprob} where we average predictions generated from a pre-trained instance of the \emph{TL} model. Even so the \emph{GP-ARX} model gives a comparable correlation coefficient to \emph{TL}, although the training set of the \emph{GP-ARX} model has no overlap with the storm events which is not the case for \emph{TL}.

@@

Figure \ref{fig:timingErr} compares the time discrepancy of predicting the storm peak. By definition, the \emph{Persistence} model will always have $\Delta t_{peak} = -1$. The \emph{GP-AR} and \emph{GP-ARX} models give better performance than the other models in terms of timing error in prediction of storm peaks.

\begin{figure}
   \centering
   \includegraphics[width=\textwidth]{Compare_pred.png}
      \caption{Comparison of OSA predictions generated by the \emph{NM}, \emph{TL} and \emph{GP-ARX} models for the storm event $9^{th}$ November $2004$ 11:00 UTC - $11^{th}$ November $2004$ 09:00 UTC}
         \label{fig:predictions}
\end{figure}

Figure \ref{fig:relprob} compares the empirical cumulative distribution functions of the absolute value of the relative errors across all storm events in the experiments.  One can see that $95\%$ of the hourly predictions have a relative error smaller than $25\%$ for \emph{GP-ARX}, $50\%$ for \emph{NM} and $60\%$ for \emph{TL}.

In figure \ref{fig:predictions} we show the OSA predictions generated by the \emph{TL}, \emph{NM} and \emph{GP-ARX} models for one event from the list of storms in \citet{Ji2012}. This storm is interesting due to its strength ($Dst_{min} = -289 \ nT$) and the presence of two distinct peaks. In this particular event the \emph{GP-ARX} model approximates the twin peaks as well as the onset and decay phases quite faithfully, contrary the \emph{NM} and \emph{TL} model predictions. The \emph{TL} model recognises the presence of twin peaks but overestimates the second one as well as the decay phase, while the \emph{NM} model is much delayed in the prediction of the initial peak and fails to approximate the time and intensity of the second peak.

As noted in section \ref{sec:inference}, \emph{Gaussian Process} models generate probabilistic predictions instead of point predictions, giving the modeller the ability to specify error bars on the predictions. In figure \ref{fig:predictionswitherrorbar} \emph{OSA} predictions along with lower and upper error bars generated by the \emph{GP-ARX} model are depicted for the event shown in figure \ref{fig:predictions}. The error bars are calculated at one standard deviation on either side of the mean prediction, using the posterior predictive distribution in equations \ref{eq:posteriormean} and \ref{eq:posteriorcov}. We remind that, since the model produces Gaussian output distributions, one standard deviation on either side of the mean corresponds to a confidence interval of $68\%$.

\begin{figure}
   \centering
   \includegraphics[width=\textwidth]{Compare_pred_err_bar.png}
      \caption{OSA predictions with error bars $\pm 1$ standard deviation, generated by the \emph{GP-ARX} model for the storm event $9^{th}$ November $2004$ 11:00 UTC - $11^{th}$ November $2004$ 09:00 UTC}
         \label{fig:predictionswitherrorbar}
\end{figure}

@@

In this paper, we proposed two \emph{Gaussian Process} auto-regressive models, \emph{GP-ARX} and \emph{GP-AR}, to generate hourly predictions of $Dst$. We compared the performance of our proposed models with the \emph{Persistence} model and six existing model benchmarks reported in \citet{Ji2012}. Predictive performance was compared on \emph{RMSE}, \emph{CC}, $\Delta Dst_{min}$ and $|\Delta t_{peak}|$ calculated on a list of 63 geomagnetic storms from \cite{Ji2012}.

Our results can be summarized as follows.
   \begin{enumerate}
      \item \emph{Persistence} model must be central in the model evaluation process in the context of \emph{one step ahead} prediction of the $Dst$ index. It is clear that the \emph{persistence} behavior in the $Dst$ values is very strong i.e. the trivial predictive model $\hat{Dst}(t) = Dst(t-1)$ gives excellent performance according to the metrics chosen. In fact it can be seen in figure \ref{sec:res} that the \emph{Persistence} model outperforms every model compared by \citet{Ji2012} in the context of OSA prediction of $Dst$. Therefore any model that is proposed to tackle the OSA prediction problem for $Dst$ should be compared to the \emph{Persistence} model and show visible gains above it.

@@

      If one compares the training sets used by the models, one can appreciate that they are able to learn robust models for prediction with relatively small training and validations sets: the training set contains 250 instances, while the validation set contains 432 instances. On the other hand the \emph{NM} model uses one year (8760 instances) of data to learn the formula outlined in \citet{balikhin:narmax} and the \emph{TL} model uses six years of data (52560 instances) for training.

      The encouraging results of Gaussian Processes illustrate the strengths of the Bayesian approach to predictive modeling. Since the GP models generate predictive distributions for test data and not just point predictions they lend themselves to the requirements of space weather prediction very well because of the need to generate error bars on predictions.
   \end{enumerate}

\subsection{Future Research Directions}

Multi-step ahead prediction should be the preferred gold standard with respect to forecasting of geomagnetic indices. Also important is the usage of more informative performance metrics for evaluating and comparing empirical predictive models.
